Efficient static analysis of Marlowe contracts

### 2020-05-27 papers/static_analysis2020/paper.tex
Reduce to 16.5 pages

\maketitle
\begin{abstract}
SMT solvers can verify properties automatically and efficiently, and they offer increasing
flexibility on the ways those properties can be described. However, the way in which
properties are described can affect the computational cost of verifying them considerably. In addition, it is hard to predict how changes on the way of describing a property
translate into changes to the computational costs.

In this paper, we discuss the lessons we have learned while implementing and optimising the static analysis functionality for Marlowe, a domain specific language for describing self-enforcing financial smart-contracts that can be deployed on a blockchain.
\end{abstract}
\section{Introduction\label{sec:Introduction}}

Thanks to static analysis, we can automatically check beforehand whether any
payments promised by a Marlowe contract can be fulfilled in every
possible execution of the contract. If a Marlowe contract has passed
the static analysis, we will have a very high assurance
that whenever the contract says we will be paid a certain amount of
money, any execution of the contract will indeed have enough money to
make such payment.

Marlowe's static analysis relies on SMT solvers, which are able to check whether a
set of constraints is satisfiable very efficiently. Thanks to state of the art
libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
high-level language. In the case of SBV, we can directly write properties as
Haskell functions, with very few restrictions on how those functions are
implemented; SBV will automatically translate those functions into a language
that SMT solvers can understand (namely, SMTLib format \cite{barrett2010smt}).

Nevertheless, static analysis can become very computationally expensive if
implemented naïvely, because it needs to consider every possible value of the
property we want to check (in our case the main property we want to check
is whether a Marlowe contract will have enough money to make all the
payments described by the contract). Indeed, high-level abstractions
are often a tradeoff with efficiency.

For example, in the case of Marlowe, contracts are typically interactive
and must deal with inputs arriving from different users in different
orders, and sometimes not arriving at all. In addition, Marlowe allows
several inputs to be grouped in a single transaction, and different
decisions may be taken depending on the content of those inputs, or
in the combinations of contents of those inputs. All these alternatives
multiply the search space of static analysis and can easily compose
exponentially and thus make the whole approach unusable in practice
even for moderately sized contracts.

This paper contributes a number of approaches that can be used when
optimising static analysis, with examples extracted from a case-study where
we applied these approaches, an approach to ensure correctness
of the optimisations that combines the usage of property-based testing
and verification using automated theorem provers, and empirical data
that measures the effect of some of the optimisations on the implementation
of our case-study.

The case study presented is that of the optimisation of Marlowe's
static analysis. During this work, the techniques applied allowed
us to reduce the analysis time of an implementation that followed
the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract would take less
than a second, and where a four-person crowdfunding contract that
when fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes using Z3. In the original implementation, we were
not able to analyse the 19 megabyte contract because the analysis
ran out of memory after half an hour.

There are many things that can be done to improve the efficiency of
static analysis. We have classified them as lightweight and heavyweight.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item \textbf{Removing unnecessary parts from the analysis}. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item \textbf{Avoiding high level abstractions}. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item \textbf{Reducing the search space by normalising parameters}. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, we can remove all but one when
implementing static analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. But these optimisations can also
translate in important reductions to execution time and memory usage, as
shown by the experiments we report in Section~\ref{sec:Measurements}.
We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing the search space by using normalised execution paths relevant
to the property}. Instead of using the search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form (or equivalence partitioning).
\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we cover a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the heavyweight optimisations described in the previous section.
In Section~\ref{sec:Measurements}, we present empirical results that
show the effects of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis.

\section{Marlowe and how its design helps static analysis and correctness\label{sec:Marlowe-intro} }

Firstly, we will briefly introduce the semantics of Marlowe, the guarantees
that are offered implicitly by the semantics, and how the design choices
facilitate static analysis and make it decidable. A more detailed
explanation of Marlowe can be found in \cite{marlowe}.

@@

Marlowe contracts are able to receive payments, store money and tokens,
ask for input from the participants, and redistribute stored money
and tokens back among participants. The decision about which and when of
these actions are carried out is determined by the contract on the basis
of the information that is available at each point in time. Participants
may correspond to either individual public keys or to tokens (\texttt{Role}s).
In turn, \texttt{Role}s may be controlled by separate independent contracts.

@@

There are 5 constructs of type \texttt{Contract}:
\begin{itemize}
\item \texttt{Close} -- signals the end of the life of a contract. Once
it is reached, all the money and tokens stored in the accounts of
the contract will be refunded to the owner of each of the respective
accounts.
\item \texttt{If} -- immediately decides how the contract must continue
from between two possibilities and depending on a given Boolean condition
that we call an \texttt{Observation}. The value of an \texttt{Observation}
can depend on previous choices made by participants, on amounts of money
and tokens stored in the contract, or other factors.
\item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
Expressions of type \texttt{Value} in Marlowe are evaluated to integer
values and they depend on information available to the contract at the
time of evaluation. For example, a \texttt{Let} construct can be used to
store the amount of money available in a particular account at a given point
in time, for use at a later stage of the contract execution.
\item \texttt{When} -- waits for an external input. The \texttt{When} construct
also specifies a timeout slot\footnote{Slots are blockchain's proxy for time,
they are added to the blockchain periodically.}: after this slot has been reached, the
\texttt{When} construct expires, and no longer accepts any input.
There are three kinds of input:
\begin{itemize}
\item \texttt{Deposit} -- Waits for a participant to deposit an amount
of money or tokens (specified as a \texttt{Value}) in an account of the contract.
\item \texttt{Choice} -- Waits for a participant to make a choice. A choice
is represented as an integer number from a set specified by the contract.
\item \texttt{Notify} -- Waits for an \texttt{Observation} to be true.
Because contracts are reactive (they cannot initiate transactions),
it is necessary for an external actor to \texttt{Notify} the contract
that the \texttt{Observation} has become true. However, this notification
does not have to be carried out by a participant, anybody can do it.
\end{itemize}
\item \texttt{Pay} -- immediately makes a payment between accounts of the
contract, or from an account of the contract to a given participant.
The amount transferred is specified as a \texttt{Value}.
\end{itemize}

@@

As we mentioned in the previous section, Marlowe
contracts are passive: their code is executed as part of the validation of
transactions that are submitted to the blockchain. These transactions
need to be submitted by participants or their representatives (e.g.
user wallets) and their validation is carried out atomically and
deterministically.

Each transaction may include a list of inputs, a set of signatures,
a slot interval\footnote{We use a slot interval because it is very difficult to
know in practice in which slot the transaction is going to be included,
and we need transactions to be deterministic, so we cannot use individual
slot numbers.}, a set of input UTxOs (incoming money and tokens), and a
set of outputs (outgoing money and tokens). For
a transaction to be valid in Marlowe, the transaction must have the
same effect for every slot within that slot range (be deterministic).
For example, if a transaction has a minimum slot number which is before
a timeout, and a maximum slot which after a timeout then the transaction
will fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

@@

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item \textbf{Classification of money and tokens into accounts} separates concerns.
A Marlowe contract will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item \textbf{Account identifiers include an account owner}. An account owner
is a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together, these properties ensure that no money or tokens
are locked in the contract forever.
\item \textbf{No negative deposits or payments}. Marlowe treats negative
amounts in deposits and payments as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully: it will still wait for a null deposit to
be made, and it will continue as if everything is correct; this way,
the expected execution of the contract is disturbed as little as possible.
\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
can accept throughout its lifetime. This limit is implied by the maximum
number of nested \texttt{When} constructs in the contract, since only one input
per \texttt{When} can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This bound prevents DoS attacks, and it makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

In this section, we briefly present and reflect on a technique that
can be used to convert a concrete implementation of a Haskell function
into a symbolic one by using the SBV library \cite{erkok2019sbv}, and to use this symbolic
implementation for static analysis. In particular, we explore this
technique in the context of the Marlowe semantics.

@@

The SBV library supports implementing Haskell functions in a way that
the same implementation can be used:
\begin{itemize}
\item With concrete parameters, as a normal Haskell function
\item With symbolic variables, so that properties can be checked for satisfiability
using an SMT solver
\item As part of QuickCheck properties, for random testing.
\end{itemize}
Parameters that can be used symbolically are wrapped in a monad called
\texttt{SBV}. Functions that depend on symbolic parameters also return
values wrapped in the \texttt{SBV} monad.

The idea is therefore to convert our implementation to use values
wrapped in the \texttt{SBV} monad, by replacing all operations on
those values with symbolic functions. In order to simplify this work,
\texttt{SBV} provides symbolic versions for many of the most common
built-in functions that are either overloaded or have similar names.

@@

We can then define a function that takes a list of transactions instead,
and we would just need to write a property that says that the result
of executing the list of transactions for a given contract does not
have any warnings (the property we are trying to verify). Then we
can pass the property to SBV and it will tell us which transaction
lists break the property.

Unfortunately, there are a couple of issues with this approach that
we review in Section~\ref{subsec:considerations}.

@@

SBV does not currently support complex custom data types. At the time
of writing, SBV has some support for 0-ary data types, but not for
data types with parameters.
It does provide symbolic versions of the \texttt{Either} type and the \texttt{Tuple} type.
Thus, it is possible to have an isomorphism between custom data types and
nested \texttt{Either} and \texttt{Tuple} type-synonyms. Our original
implementation automatically generates conversion functions between these
by using Template Haskell \cite{sheard2002template}. By using this technique the symbolic
implementation can remain similar to the concrete one.

For example, the following data structure:

@@

But this approach cannot address recursive datatypes, let alone mutually
recursive datatypes. And the \texttt{Contract} type is mutually recursive.

Even if we could represent mutually recursive datatypes, it would
not be feasible to directly translate the semantics, because a more
general limitation of SMT solvers is that they cannot, in general,
deal with arbitrarily sized computations. In other words, if termination
of a symbolic function is not bounded, or is only bounded by a symbolic
parameter, SMT solvers will often fail to terminate when trying to validate
a property, since doing so often requires a proof by induction. We
discuss how to address this problem in Section~\ref{subsec:Adapting-the-semantics}.

@@

In order to guarantee termination of the SMT validation, we need a
concrete bound for the property that we want to validate.

Related work often addresses this problem by manually establishing
a bound on the amount of computation, e.g: limiting the number of
computation steps analysed, the number of times loops are unrolled
\cite{clarke2004tool,gulwani2008program,jackson2000finding}.

In the case of Marlowe, we have natural bounds that we can use because,
given a concrete contract, we can easily infer:
\begin{itemize}
\item The maximum number of inputs that can have an effect on the contract
\item The maximum number of transactions that can have an effect on the
contract
\item All of the account, choice, and \texttt{Let} identifiers that will
be used in the contract
\item The number of participants that will participate in the contract
\end{itemize}
From this data, we can also deduce an upper bound for:
\begin{itemize}
\item The number of times that
\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} may
be called.
\item The number of accounts that the contract will use, and
the number of elements there may be in each of the associative maps
that comprise the \texttt{State} of the contract at any given point
of the execution.
\end{itemize}
If we keep the \texttt{Contract} parameter concrete,
we can use it to bound the execution, because every call to \texttt{reduceContractStep}
will either make no progress or remove one of the constructs, with
the exception of \texttt{Close}. But for any particular \texttt{Contract},
we know we only need to call \texttt{reduceContractStep} at most as many times as
the maximum number of accounts there can possibly be, since, every time we call
it, it will refund one of the accounts.

Thus, the symbolic transaction processing function becomes:

@@

We get around this problem by using a continuation style. Instead
of returning the \texttt{TransactionOutput} directly, we take a continuation
function that takes the concrete \texttt{Contract} and the symbolic version
of \texttt{TransactionOutput} without the \texttt{Contract}.

@@

In practice, we also include some extra information about bounds,
and we make some other parts of \texttt{TransactionOutput} concrete,
in addition to the \texttt{Contract}.

@@

In this section, we explain, in more detail, the techniques that we mentioned
in the introduction, and we illustrate those techniques with
examples from the static analysis implementation of Marlowe.

@@

Some of the modifications that we can apply to a static analysis implementation
are localised, and they do not fundamentally affect the overall structure
of the implementation. This means that the implementation can remain close
to the original semantics and model that we are
trying to analyze, as well as making changes less likely to introduce
errors than a more serious restructuring.

@@

This is probably the most straightforward optimization to implement.
When we use the same or similar code for both the analysis and the
implementation, we may end up including code that is not relevant
to the analysis.

In the case of Marlowe, this was the case of the \texttt{Close} construct.
The \texttt{Close} construct refunds all the money and tokens remaining
in the accounts. At the time of analysis, we do not know how much
is available in each account, because it may depend on the value of
a choice made by the user in runtime. For that reason, the generated
constraints must include comparisons for every account. This makes
the implementation of the \texttt{Close} construct very expensive
to execute symbolically.

As it turns out, we do not need to implement the semantics of the
\texttt{Close} construct at all, because it is impossible for the
\texttt{Close} construct to produce a failed payment
(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} (last visited on 04/04/2020)} using Isabelle \cite{nipkow2002isabelle}): it only pays whatever there is available.
So we can safely remove it from the analysis.

@@

High level libraries like SBV, and even standards like SMTLib, support
the use and construction of high level abstractions,
e.g: custom data-types, list, sets\dots{} Unfortunately, even though
high level abstractions make reasoning about code easier, they often
prevent certain optimisations, since they abstract out aspects of
the implementation that in our particular case may be concrete.

For example, in the case of Marlowe's static analysis, we initially
implemented a \emph{symbolic} associative map primitive with the only limitation
that it needed a concrete bound in the number of elements. This is
straightforward to realise using the symbolic implementation of list
and tuple, both provided by SBV. However, this can be very costly:
searching a list symbolically requires constraints that check
the key of every element in the list up to the maximum capacity of
the associative map, because it is not known at the time of symbolic
execution how many elements the list has.

However, in Marlowe we know the values of all the keys that we are
going to use in maps, because the contract is concrete, and only \texttt{Account},
\texttt{Choice}, and \texttt{Let} identifiers that are mentioned in
the contract will ever make it into the \texttt{State}. So we do not
need keys of the associative map to be symbolic, we can use a concrete
associative map with symbolic values.

Nevertheless, using a partially concrete map prevents us from using
it as the result of a function that takes symbolic parameters, so
we will need to use a continuation style insead. An alternative is
to assign numbers to each possible key, and store them in a symbolic
list, this will at least prevent the search from having to compare
a lookup with every key, since we will know in which position of the
list each value is stored.

@@

The higher the number of degrees of freedom of the input, the larger
the search space, and the higher the load we are putting on the SMT
solver. In many cases, we do not need to explore the whole search
space to check satisfiability of a property. In particular, we may
be able to prove that some sets of inputs are equivalent in terms
of the property whose satisfiability we want to test.

For example, in the case of Marlowe's static analysis,
Marlowe allows several inputs to be combined into a single transaction.
This functionality is important because each transaction requires
the issuer to pay fees, so combining several inputs in a single transaction
can translate into important savings. On the other hand, it also means
that static analysis must consider many more possibilities, since
the number of ways of partitioning inputs in transactions grows exponentially
in the number of inputs.

@@

This optimization reduces the search space considerably, but transactions
may still have no inputs, so there are still many ways of distributing
the inputs in transactiongs. We explain how we reduced
the search space even further in Section~\ref{subsec:Heavyweight-modifications}.

@@

When optimising, if our solution is a local minimum, small changes
to the parameters will not grant any improvement to the result.
For that reason, in this section, we explore ways of optimizing that
may imply considerable rewriting of our properties, constraints, and
static analysis implementation in general.

@@

One approach we can use that helps improve static analysis efficiency
is to focus on the property we want to test instead of trying to make
it as close as possible to the full implementation
of the system. We do not even need to think about the representation
of a counterexample (we will discuss that in the next section), but
only in what are the conditions for the property to be false.

Unfortunately, having static analysis be very different from implementation
makes the approach much more error prone, and in making an argument
for the trustworthiness of the analysis there are many more assumptions
that need to be made and reasoned about. In Section~\ref{sec:Testing},
we explore ways of mitigating this problem.

For example, the main property we want to check is whether there is
any possible execution that produces a failed payment. Thus, we look
at executions instead of looking at inputs. The most complicated construct
in terms of executions is the \texttt{When} construct, since it allows
for two transactions that are separate in time to have different effects
depending on when they are issued, all other constructs will get resolved
instantly in one way or another. But, without loss of generality,
we can structure them as Figure~\ref{fig:Distribution-of-transactions}
shows: we can conceptually break the contract tree into subtrees,
where each subtree has a \texttt{When} construct as its root, with
the exception of the subtree at the root of the original tree.

\begin{figure}
\includegraphics[width=1\textwidth]{figures/transactions}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=0.45\textwidth]{figures/equivalence_property}
\par\end{centering}
\caption{Property based testing for equivalence\label{fig:Property-based-testing-for-equivalence}}
\end{figure}

@@

\begin{table}[]
\centering
\scalebox{0.8}{

>>>>

\maketitle
\begin{abstract}
SMT solvers can verify properties automatically and efficiently, and they offer increasing
flexibility on the ways those properties can be described. But it is hard to predict how those ways of describing the properties affect the computational cost of verifying them.

In this paper, we discuss what we learned while implementing and optimising the static analysis for Marlowe, a domain specific language for self-enforcing financial smart-contracts that can be deployed on a blockchain.
\end{abstract}
\section{Introduction\label{sec:Introduction}}

Thanks to static analysis, we can automatically check beforehand whether any
payments promised by a Marlowe contract can be fulfilled in every
possible execution of the contract. If a Marlowe contract has passed
the static analysis, we will have a very high assurance
that whenever the contract says it will make a payment, the contract will
indeed have enough money available.

Marlowe's static analysis relies on SMT solvers, which are able to check efficiently
whether a set of constraints is satisfiable. Thanks to state of the art
libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
high level language. In the case of SBV, we can write properties as
Haskell functions, with few restrictions on how those functions are
implemented; SBV automatically translates those functions to SMTLib format \cite{barrett2010smt}, a language many SMT solvers understand.

Nevertheless, static analysis can become very computationally expensive if
implemented naïvely, because it is an NP-complete problem in the general case.
In our case the main property is whether a contract will have enough money for all payments. Indeed, high level abstractions are often a tradeoff with efficiency.

% For example, in the case of Marlowe, contracts are typically interactive
% and must deal with inputs arriving from different users in different
% orders, and sometimes not arriving at all. In addition, Marlowe allows
% several inputs to be grouped in a single transaction, and different
% decisions may be taken depending on the content of those inputs, or
% in the combinations of contents of those inputs. All these alternatives
% multiply the search space of static analysis and can easily compose
% exponentially and thus make the whole approach unusable in practice
% even for moderately sized contracts.

This paper contributes a number of approaches that can be used when
optimising static analysis, with examples extracted from a case-study where
we applied these approaches, an approach to ensure correctness
of the optimisations that combines the usage of property-based testing
and verification using automated theorem provers, and empirical data
that measures the effect of some of the optimisations on the implementation
of our case-study: the optimisation of Marlowe's static analysis.

The techniques described helped us reduce the analysis time of an implementation
that followed the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract takes less
than a second to analyse, and where a four-person crowdfunding contract that
fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes.

We have classified optimisation techniques as lightweight and heavyweight.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item \textbf{Removing unnecessary parts from the analysis}. If they do not affect
the property that is being verified then we can just remove them.
\item \textbf{Avoiding high level abstractions}. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that may not be necessary.
\item \textbf{Reducing the search space by normalising parameters}. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, we can remove all but one for analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. But these optimisations can also
translate in important reductions to execution time and memory usage, as
shown by the experiments we report in Section~\ref{sec:Measurements}.
We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing the search space by using normalised execution paths relevant
to the property}. Instead of using the search space to model inputs, we
use it to model possible executions, and we only
represent each equivalence class of executions once, i.e:
we represent them in a normal form.
\item \textbf{Minimizing the representation of inputs and outputs}. We encode inputs
and outputs as concisely as possible, discarding inferable information.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we cover a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we illustrate the use of property based testing on the implementation with
heavyweight optimisations. Finally, we present empirical results that
show the effect of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis (Section~\ref{sec:Measurements}).

\section{Marlowe design and semantics\label{sec:Marlowe-intro} }

Firstly, we introduce the semantics of Marlowe, the guarantees
that are offered implicitly by the semantics, and how the design choices
facilitate static analysis and make it decidable. A more detailed
explanation can be found in \cite{marlowe}.

@@

Marlowe contracts are able to receive payments, store money and tokens,
ask for input from its participants, and redistribute stored money
and tokens among participants. A contract determines when and which of
these actions may be carried out. Participants may correspond to either
individual public keys or to tokens (\texttt{Role}s). In turn, \texttt{Role}s
may be controlled by other contracts.

@@

There are 5 constructs of type \texttt{Contract}:
\begin{itemize}
\item \texttt{Close} -- signals the end of the life of a contract. Once
it is reached, all the money and tokens stored in the accounts of
the contract will be refunded to the owner of each of the respective
accounts.
\item \texttt{If} -- immediately decides how the contract must continue
from between two possibilities, depending on a given Boolean condition
that we call an \texttt{Observation}. An \texttt{Observation}
may depend on previous choices by participants, on amounts of money
and tokens in the contract, or other factors.
\item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
Expressions of type \texttt{Value} in Marlowe are evaluated to integer
values and they depend on information available to the contract at the
time of evaluation. For example, a \texttt{Let} construct record
the amount of money in an account at a point in time.
\item \texttt{When} -- waits for an external input. The \texttt{When} construct
also specifies a timeout slot\footnote{Slots are blockchain's proxy for time,
they are added to the blockchain periodically.}: after this slot has been reached, the
\texttt{When} construct expires, and no longer accepts any input.
There are three kinds of input:
\begin{itemize}
\item \texttt{Deposit} -- Waits for a participant to deposit an amount
of money or tokens (specified as a \texttt{Value}) in an account of the contract.
\item \texttt{Choice} -- Waits for a participant to make a choice. A choice
is represented as an integer number from a set specified by the contract.
\item \texttt{Notify} -- Waits for an \texttt{Observation} to be true.
Because contracts are reactive (they cannot initiate transactions),
it is necessary for an external actor (not necessarily a participant)
to \texttt{Notify} the contract.
\end{itemize}
\item \texttt{Pay} -- immediately makes a payment between accounts of the
contract, or from an account of the contract to a given participant.
The amount transferred is specified as a \texttt{Value}.
\end{itemize}

@@

As we mentioned in the previous section, Marlowe
contracts are passive: their code is executed as part of the validation of
transactions that are submitted to the blockchain. Transactions
need to be submitted by participants or their representatives (e.g.
user wallets) and validation is atomic and deterministic.

Each transaction may include a list of inputs, a set of signatures,
a slot interval, a set of input UTxOs (incoming money and tokens), and a
set of outputs (outgoing money and tokens). We use a slot interval because
it is very difficult to know the exact slot in which the transaction will
be included in practice. For a transaction to be valid in Marlowe, the
transaction must have the same effect for every slot within that slot
interval (be deterministic). For example, if a transaction has a minimum
slot number lower than timeout, and a maximum slot that is greater, then the transaction will fail with the error \texttt{AmbiguousSlotInterval}.

@@

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item \textbf{Classification of money and tokens into accounts} separates concerns.
A Marlowe contract will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But a payment
for more than there is will not fail, it will pay as much as
is available, in order to remain as close as possible
to the original intention of the contract.
\item \textbf{Account identifiers include an account owner}. An account owner
is a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together, these properties ensure that no money or tokens
are locked in the contract forever.
\item \textbf{No negative deposits or payments}. Marlowe treats negative
amounts in deposits and payments as zero. At the same time,
if there happens to be a request for a deposit with a negative amount,
the contract will still wait for a null deposit to
be made, and it will continue as if everything is correct. This way execution
of the contract is disturbed as little as possible.
\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
can accept throughout its lifetime. This limit is implied by the path with
maximum number of nested \texttt{When} constructs in the contract, since only
one input per \texttt{When} can be accepted. At the same time, transactions
with no effect on the contract are invalid, thus there is
a limit on the maximum number of transactions a contract can accept
throughout its life too. This bound prevents DoS attacks, and it makes static analysis
easier. We discuss further in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

In this section, we briefly present and reflect on a technique that
can be used to convert a concrete implementation of a Haskell function
into a symbolic one by using the SBV library \cite{erkok2019sbv}, and to use this symbolic
implementation for static analysis. In particular, we explore this
technique in the context of the Marlowe.

@@

The SBV library supports implementing Haskell functions in a way that
the same implementation can be used:
\begin{itemize}
\item With concrete parameters, as a normal Haskell function.
\item With symbolic parameters, so that properties can be checked for satisfiability
using an SMT solver.
\item As part of QuickCheck properties, for random testing.
\end{itemize}
Parameters that can be used symbolically are wrapped in a monad called
\texttt{SBV}. Values that depend on symbolic values must also be wrapped in the \texttt{SBV} monad.

@@

We just need a function \texttt{playTrace} that takes a list of
transactions and calls \texttt{computeTransaction} for each. 
We can then write our property to state that the output of
\texttt{playTrace} does not have any failed payments (or other warnings). We
can ask SBV to find an input transaction list that breaks the property.
However, there are a couple of issues with this approach,
we review them in Section~\ref{subsec:considerations}.

@@

At the time of writing, SBV does not fully support complex custom data types,
but it provides symbolic versions for \texttt{Either} and \texttt{Tuple} types.
Our original implementation of Marlowe's static analysis overcomes this limitation
by generating conversion functions using using Template Haskell
\cite{sheard2002template}. This allows static analysis to remain similar to the semantics. For example, the following data structure:

@@

But this approach cannot address recursive datatypes, let alone mutually
recursive datatypes. And the \texttt{Contract} definition uses mutual recursion.

Even if we could use symbolic corecursive datatypes, SMT solvers have another
general limitation: if termination of a function is not bounded
by a concrete value, SMT solvers may not terminate when determining
the satisfiability of a property about the function.
We discuss how to address this in Section~\ref{subsec:Adapting-the-semantics}.

@@

In order to guarantee termination of the analysis, we need a
concrete bound. Related work often addresses this problem by
manually establishing an artificial bound on the amount of computation,
e.g: limiting the number of computation steps analysed,
the number of times loops are unrolled \cite{clarke2004tool,gulwani2008program,jackson2000finding}.

Marlowe has natural bounds, given a concrete contract, we can infer:
\begin{itemize}
\item The maximum number of inputs that can have an effect on the contract
\item The maximum number of transactions that can have an effect on the
contract
\item All of the account, choice, and \texttt{Let} identifiers that will
be used in the contract
\item The number of participants that will participate in the contract
\end{itemize}
From this data, we can also deduce an upper bound for:
\begin{itemize}
\item The number of times that
\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} may
be called.
\item The number of accounts that the contract will use, and
the number of elements there may be in each of the associative maps
that comprise the \texttt{State} of the contract at any given point
of the execution.
\end{itemize}
Marlowe \texttt{Contract}s are finite, and every call to
\texttt{reduceContractStep} will either make no progress or
remove one of the constructs, with the exception of \texttt{Close}.
In the case of \texttt{Close}, every call to \texttt{reduceContractStep} refunds
one account, and the number of accounts is also bounded, since each
needs to be mentioned in the \texttt{Contract}. Thus, the symbolic transaction
processing function becomes:

@@

We get around this by using a continuation style. Instead
of returning the \texttt{TransactionOutput}, we take a continuation
function that takes the concrete \texttt{Contract} and the symbolic version
of \texttt{TransactionOutput} without the \texttt{Contract}.

@@

In practice, we also include some extra information about bounds,
and we make some other parts of \texttt{TransactionOutput} concrete.

@@

In this section, we explain the optimisation techniques in more detail,
and we illustrate them with examples of their application to Marlowe's static analysis.

@@

Lightweight modifications are local, which means it is less likely that we
will introduce reasoning errors when implementing them.

@@

When we use the same or similar code for both the analysis and the
implementation, we may end up including code that is not relevant
to the analysis.

In the case of Marlowe, this was the case of the \texttt{Close} construct.
The \texttt{Close} construct refunds all the money and tokens remaining
in the accounts. The number of times that \texttt{reduceContractStep} needs
to be called depends on how many accounts have money left, and because
this information is symbolic, there are many potential ways in which execution
can unfold. All these paths need to be represented as constraints, which
makes analysing \texttt{Close} very costly.

Fortunately, as it turns out, we do not need to analyse \texttt{Close} at all.
Because it is impossible for the \texttt{Close} construct to produce a failed payment
or any other warning (we have proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} (last visited on 04/04/2020)} using Isabelle \cite{nipkow2002isabelle}). \texttt{Close} only pays as much as available,
so we can safely remove it from the analysis.

@@

High level libraries like SBV, and even standards like SMTLib, support
the use and construction of high level abstractions,
e.g: custom data-types, list, sets\dots{} Unfortunately, even though
high level abstractions aid reasoning about code, they often
prevent optimisations, since they abstract out aspects that
in our particular case may be concrete.

For example, in the case of Marlowe's static analysis, we initially
implemented a \emph{symbolic} associative map primitive with the only limitation
that it needed a concrete bound in the number of elements. This is
straightforward to realise using the symbolic implementation of list
and tuple, both provided by SBV. However, because we assumed keys
were symbolic, looking up a single element required constraints that
compared the element with every key up to the maximum capacity of
the associative map.

Nevertheless, in Marlowe we know the values of all the keys that we are
going to use in maps, because the contract is concrete, and only \texttt{Account},
\texttt{Choice}, and \texttt{Let} identifiers that are mentioned in
the contract will ever make it into the \texttt{State}. So we do not
need keys of the associative map to be symbolic, we can use a concrete
associative map with symbolic values.

@@

The higher the number of degrees of freedom of the input, the larger
the search space, and the higher the load we are putting on the SMT
solver. But, if two or more different inputs have the same effect on the
property we only need to include one of them.

For example, in the case of Marlowe's static analysis,
Marlowe allows several inputs to be combined into a single transaction.
This functionality is important because each transaction requires
the issuer to pay fees. On the other hand, it also means
that static analysis must consider many more possibilities, since
the number of ways of partitioning inputs is exponential
in the number of inputs.

@@

This optimization reduces the search space, but transactions may still
have either one or zero inputs, so there are still many ways of distributing
the inputs in transactions. We explain how we reduced
the search space further in Section~\ref{subsec:Heavyweight-modifications}.

@@

When optimising, if our solution is a local minimum, small changes
to the parameters will not grant any improvement to the result.
For that reason, in this section, we explore ways of optimising that
may imply considerable rewriting of our properties, constraints, and
static analysis implementation in general.

Unfortunately, not following the concrete implementation closely is
much more error prone, since there are many
more assumptions that we need to make and reason about.
In Section~\ref{sec:Testing}, we explore ways of mitigating this issue.

@@

Instead of modelling the execution symbolically, we can focus on modelling
the property. We do not even need to consider the
representation of the counterexamples (we will discuss that in the next section),
but only in what are the conditions for the property to be false.

For example, the main property we want to check is whether there is
any possible execution that produces a failed payment. Thus, we only
write constraints for executions in which this can happen instead of
modelling all possibilities.
The most complicated construct in terms of execution is the \texttt{When}
construct, since it allows for transactions that are separate in time
to have different effects depending on when they are issued, all other
constructs will get resolved atomically in one way or another. Without
loss of generality, we can structure possible executions as shown in
Figure~\ref{fig:Distribution-of-transactions}: we can conceptually break
the contract tree into subtrees, where each subtree has a \texttt{When}
construct as its root, with the exception of the subtree at the root
of the original tree.

\begin{figure}
\center
\includegraphics[width=0.7\textwidth]{figures/transactions}

@@

@@
